---
title: "AcqFac power analysis"
author: "morgan moyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lme4)
library(simr)
library(dplyr)

# theme_set(theme_bw())
cbPalette <- c("#56B4E9", "#D55E00", "#009E73","#999999", "#E69F00")
this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)
```

# Generate fake data

Assumptions:

40 participants

8 conditions (2×2×2)

One trial per condition per participant (can be adjusted)

Moderate effect sizes

```{r}
set.seed(123)

# Factors
n_subj <- 100
n_trials <- 15  # per condition ~ must be sufficiently high for RE structure
levels_Verb <- c("know", "think")
levels_Complement <- c("that", "if")
levels_RootForce <- c("declarative", "interrogative")

# Full factorial design
design <- expand.grid(
  ID = factor(1:n_subj),
  Verb = factor(levels_Verb),
  Complement = factor(levels_Complement),
  RootForce = factor(levels_RootForce)
)

# Repeat trials per condition per participant
design <- design[rep(seq_len(nrow(design)), each = n_trials), ]

# Create Item variable: unique per condition
design$Item <- interaction(design$Verb, design$Complement, design$RootForce)

# Step 1: Simulate random intercepts for subjects
subject_rint <- rnorm(n_subj, mean = 0, sd = 0.5)
design$subject_intercept <- subject_rint[as.numeric(design$ID)]

# Step 2: Simulate random intercepts for items
item_levels <- levels(design$Item)  # ensure consistent mapping
item_rint <- rnorm(length(item_levels), mean = 0, sd = 0.3)
design$item_intercept <- item_rint[as.numeric(design$Item)]

# Step 3: Simulate fixed effects + random intercepts
design$linpred <- with(design,
  -0.2 +                            # Intercept (log-odds scale)
    0.5 * (Verb == "know") +
    0.4 * (Complement == "that") +
    0.3 * (RootForce == "declarative") +
    0.6 * (Verb == "know" & Complement == "that") +
    subject_intercept + item_intercept
)

# Step 4: Convert to binary outcome
design$choice_know <- rbinom(nrow(design), size = 1, prob = plogis(design$linpred))

```


# Fit the logit model

```{r, message = FALSE, warning = FALSE}

model <- glmer(choice_know ~ Verb * Complement * RootForce + 
               (1 | ID) + 
               (1 | Item), 
               data = design, 
               family = binomial)

# Full RE model has convergence issues
# model <- glmer(choice_know ~ Verb * Complement * RootForce + 
#                  (1 + Verb * Complement * RootForce|ID) + 
#                  (1|Item),
#                data = design, 
#                family = binomial)

```

# Power analysis

If there really is a three-way interaction in the population, how often would my model detect it as statistically significant (p < 0.05) with my sample size?

To answer that, simr uses simulation to repeatedly simulate new datasets based on your fitted model, re-fits the model on each one, and checks how often the three-way interaction is significant.


Here's how simr works:

Simulates 100 new datasets using your fitted model’s structure (same fixed and random effects, same number of subjects and trials).

For each dataset:

Fits the full model (with the three-way interaction)

Fits a reduced model (without the three-way interaction)

Compares them using a likelihood ratio test

Counts how many times the p-value is < 0.05.

Returns the proportion of successful detections = estimated power.

```{r, message = FALSE, warning = FALSE}
result <- powerSim(model, fixed("Verb:Complement:RootForce", "lr"), nsim = 400)
saveRDS(result, "powerSim_output.rds")
result
```

Based on 100 simulations, the model detected a statistically significant effect (p < .05) in 31% of cases.

This means you currently have 31% power to detect the 3-way interaction — well below the conventional 80% threshold.

The 95% confidence interval on this estimate is from 22.13% to 41.03%. That reflects simulation uncertainty.

# Estimate sample size

```{r, message = FALSE, warning = FALSE}

# Extend model for more participants
model_extended <- extend(model, along = "ID", n = 500)
saveRDS(model_extended, "model_extended_output.rds")
model_extended

# Power curve for 3-way interaction
pc <- powerCurve(model_extended, fixed("Verb:Complement:RootForce", "lr"), along = "ID", nsim = 50)
plot(pc)
```



